A Deep Learning Approach for Gait Event Detection from a Single Shank-Worn IMU: Validation in Healthy and Neurological Cohorts

### 2022-03-18 paper/MDPI_template/article.tex
Extended discussion. Add figure to explain TCN arch

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}, where the beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or passive, retro-reflective marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is more and more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

In this study, we further these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
\subsection{Data Collection}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that were assumed not to affect mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}. For some recordings, the sampling frequency of the IMUs was erroneously set at 100 Hz, and therefore data from these trials were upsampled before further analysis.

\subsection{Data Pre-processing \label{subsec:data_preprocessing}}
\subsubsection{Marker Data}
The reflective marker data were first used to determine start and end of each trial. The start of the trial was defined as the time for which the first toe marker (left or right) crossed the \emph{virtual} starting line. The end was defined as the time for which the last heel marker crossed the end line. For both marker and IMU systems, data were then cropped from start to the end of the trial.

@@

\subsubsection{IMU Data}
The idea behind the deep learning approach was that a \emph{model} was trained to predict the likelihood of an IC and FC, given accelerometer and gyroscope data from a single IMU. In other words, the model was iteratively trained to learn a mapping $h_{\mathbf{W}}(\mathbf{X}): \mathbf{X} \rightarrow \mathbf{y}$, where $h_{\mathbf{W}}$ was also referred to as the \emph{hypothesis} parameterized by the weights, collectively denoted by $\mathbf{W}$, $\mathbf{X}$ was the matrix of IMU data, and $\mathbf{y}$ was the vector of annotated events.

Here, the matrix $\mathbf{X}$ consisted of the IMU data, arranged as:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}
	\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[1] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}
where $\mathbf{x}_{d} = \begin{bmatrix}x_{d}[1] & x_{d}[2] & \cdots & x_{d}[n] & \cdots & x_{d}[N]\end{bmatrix}^{\mathrm{T}}$, with $d = 1, \cdots, D$ referred to the $d$-th channel, and $n = 1, \cdots, N$ referred to the $n$-th sample (or discrete time step).

Accordingly, the vector $\mathbf{y}$ consisted of the annotated events, arranged as:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{1}[1]\\
		y_{1}[2]\\
		\vdots\\
		y_{1}[N]
	\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{2}[1]\\
		y_{2}[2]\\
		\vdots\\
		y_{2}[N]
	\end{bmatrix}, \quad y_{i}[n] \in [0, 1]
	\label{eqn:train_labels_single_instace}
\end{equation}

The data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}).

@@

\subsection{Model}
\subsubsection{Model Architecture}
The basic architecture for the deep learning model was a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of repeating blocks of dilated convolutions (CONV) \cite{YuKoltun2016,Bai2018} that were followed by batch normalization (BN) \cite{Ioffe2015}, rectified linear unit (ReLU) activation, and dropout (DropOut) \cite{Srivastava2014}. For each given dilation factor, the sequence of layers CONV-BN-ReLU-DropOut was repeated twice \cite{Remy2020}. Dilation factors were given as a sequence of increasing powers of 2, e.g., $\left\{1, 2, 4, 8\right\}$ \cite{VanDenOord2016,YuKoltun2016,Bai2018}. These repeating blocks were followed by a fully-connected (dense) layer with sigmoid activation, and outputs were predicted separately for ICs and FCs. The mean squared error (MSE) was used as loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

\subsubsection{Hyperparameter Optimization}
In order to find the best model architecture, hyperparameter tuning was perfomed using KerasTuner \cite{OMalley2019}. Here, the number of filters, the kernel size, and the maximum dilation factor (Table \ref{tab:hyperparameter_optimization}) were optimized for using a random search strategy \cite{Bergstra2012}.

@@

The model architecture that resulted from the hyperparameter optimization was then trained on the combined set of training and validation data. The trained model was used to predict occurrence of gait events from the hold-out test set.

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived gait parameters.
\subsubsection{Overall detection performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated evens were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many event that were detected, were actually not annotated (false positives, FP). From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} \nonumber
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} \nonumber
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \cdot \textrm{precision}}{\textrm{recall} + \textrm{precision}}  \nonumber
	\label{eqn:f1_score}
\end{equation}

\subsubsection{Time Error}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), the time error between the annotated and detect gait event was defined as:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Stride Parameters}
For those trials for which all gait events were detected, and no spurious events were detected, the stride time, stance time and swing were calculated. Stride was the time between two successive ICs of the same foot. Stance time was the time between FC and the preceding IC of the same foot. Swing time was the time between the IC following the last FC of the same foot.

@@

\subsection{Overall detection performance}
The performance of detecting initial contacts and final contacts was objectively quantified by the number of annotated events that were detected (true positives, TP), the number of annotated events that were not detected (false negatives, FN), and the number of detected events that were not annotated (false positives, FP). From these, recall, precision and F$_{1}$ score were calculated (Table \ref{tab:overall_detection_performance}).
\begin{table}[H]
	\caption{Overall detection performance for initial contacts and final contacts as quantified by recall, precision and F$_{1}$ score.\label{tab:overall_detection_performance}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{r|CCCCCC|CCCCCC}
			\toprule
			 & \multicolumn{6}{c}{\textbf{Initial contacts}}	& \multicolumn{6}{c}{\textbf{Final contacts}}\\
			\textbf{Tracked point}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision}	& \textbf{F1}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision} 	& \textbf{F1}\\
			\midrule
			left ankle		& 624	& 19	& 5		& 97\%	& 99\%	& 98\%		& 606	& 32	& 10	& 95\%	& 98\%	& 97\%\\
			right ankle		& 599	& 42	& 8 	& 93\%	& 99\%	& 96\%		& 614	& 17	& 12	& 97\%	& 98\%	& 98\%\\
			left shank		& 605	& 38	& 15	& 94\%	& 98\%	& 96\%		& 585	& 53	& 18	& 92\%	& 97\%	& 94\%\\
			right shank		& 603	& 36	& 15 	& 94\%	& 98\%	& 96\%		& 595	& 30	& 9		& 95\%	& 99\%	& 97\%\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{TP: true positives, FN: false negatives, FP: false positives, F1: F$_{1}$ score.}}
\end{table}

@@

\subsection{Time Error}
\begin{figure}[H]
	\centering
	\includegraphics[width=13.5 cm]{fig/box_plots_tracked_points_with_outliers.png}
	\caption{Time errors for initial (left) and final (right) contacts detection, for each of the different tracked points.\label{fig:time_error_box_plots}}
\end{figure}

Time errors for ICs and FCs are visually depicted using boxplots for each tracked point. Data were not normally distributed, thus Wilcoxon signed-rank tests were used to evaluate whether the median time error was zero.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CC|CC}
		\toprule
		 & \multicolumn{2}{c}{\textbf{Initial contacts}} & \multicolumn{2}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & \textbf{median} & \textbf{IQR}\\
		 & s & s & s & s \\
		\midrule
		left ankle & 0.000 & 0.020 & 0.000 & 0.010\\
		right ankle & 0.000 & 0.020 & -0.005 & 0.015\\
		left shank & -0.005 & 0.020 & -0.005 & 0.020\\
		right shank & -0.003 & 0.020 & -0.005 & 0.020\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range.}}
\end{table}

\subsection{Gait parameters}
For those trials for which all gait events were correctly detected (and no spurious events were detected, i.e., no false positives), stride time, stance time and swing time were calculated. The mean difference and the limits of agreement between the NN-derived gait parameters and the marker-based annotations were calculated.

For all stride-specific gait parameters, and for all tracked points, the mean difference compared to the marker-based reference values was close to zero, i.e., the maxmum mean difference was 0.003 s. Furthermore, for all gait parameters and for all tracked points the limits of agreement, based on a 95\% confidence interval were distributed around zero with the overall span between -0.049 s and 0.051 s.
\begin{table}[H] 
	\caption{Time agreement between the stride-specific parameters. \label{tab:stride_parameters}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{rrCC}
		\toprule
		\textbf{Tracked point} & \textbf{Parameters} & \textbf{Mean difference} & \textbf{Limits of Agreement}\\
		 & & ms & (ms, ms)\\
		\midrule
		left ankle & stride time & 0.001 & (-0.035, 0.036)\\
		 & stance time & 0.002 & (-0.039, 0.042)\\
		 & swing time & -0.001 & (-0.045, 0.043)\\
		\midrule
		right ankle & stride time & 0.000 & (-0.039, 0.040)\\
		 & stance time & -0.002 & (-0.048, 0.044)\\
		 & swing time & 0.003 & (-0.046, 0.051)\\
		\midrule
		left shank & stride time & 0.001 & (-0.039, 0.041)\\
		& stance time & 0.002 & (-0.043, 0.046)\\
		& swing time & -0.001 & (-0.049, 0.047)\\
		\midrule
		right shank & stride time & -0.000 & (-0.031, 0.031)\\
		& stance time & 0.002 & (-0.046, 0.049)\\
		& swing time & -0.002 & (-0.049, 0.046)\\
		\bottomrule
	\end{tabularx}
	\noindent{}
\end{table}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction \label{sec:Introduction}}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}, where the beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or passive, retro-reflective marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is more and more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

In this study, we further these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods \label{sec:Materials_and_Methods}}
\subsection{Data Collection \label{subsec:Data_Collection}}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that were assumed not to affect mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}. For some recordings, the sampling frequency of the IMUs was mistakenly set at 100 Hz, and therefore data from these trials were upsampled before further analysis.

\subsection{Data Pre-processing \label{subsec:Data_Pre-processing}}
\subsubsection{Marker Data \label{subsubsec:Marker_Data}}
The reflective marker data were first used to determine start and end of each trial. The start of the trial was defined as the time for which the first toe marker (left or right) crossed the \emph{virtual} starting line. The end was defined as the time for which the last heel marker crossed the end line. For both marker and IMU systems, data were then cropped from start to the end of the trial.

@@

\subsubsection{IMU Data \label{subsubsec:IMU_Data}}
The idea behind the deep learning approach was that a \emph{model} was trained to predict the likelihood of an IC and FC, given accelerometer and gyroscope data from a single IMU. The data from a single sensor channel, e.g., the acceleration in forward direction, or the angular velocity about the vertical direction, was denoted by $\mathbf{x}_{d} = \begin{bmatrix}
x_{d}[1] & x_{d}[2] & \cdots & x_{d}[N]
\end{bmatrix}^{\mathrm{T}}$, with $d$ referring to the $d$th sensor channel (i.e., $d = 1, \cdots, D$) and $n$ referring to the $n$th sample or time step (i.e., $n = 1, \cdots, N$). Data from all $D$ sensor channels together was then denoted by:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}
		\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[1] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}

Likewise, the \emph{labels} were denoted by:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{\mathrm{IC}}[1]\\
		y_{\mathrm{IC}}[2]\\
		\vdots\\
		y_{\mathrm{IC}}[N]
		\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{\mathrm{FC}}[1]\\
		y_{\mathrm{FC}}[2]\\
		\vdots\\
		y_{\mathrm{FC}}[N]
		\end{bmatrix}, \quad y_{\mathrm{IC}/\mathrm{FC}}[n] \in [0, 1]
	\label{eqn:train_labels_single_instance}
\end{equation}

The model was then iteratively trained to learn a mapping $h_{\mathbf{\Theta}}(\mathbf{X}): \mathbf{X} \rightarrow \mathbf{y}$, where $h_{\mathbf{\Theta}}$ was also referred to as the \emph{hypothesis} which was parameterized by the weights and biases, collectively denoted by $\mathbf{\Theta}$.

All participant data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}).

@@

\subsection{Model \label{subsec:Model}}
\subsubsection{Model Architecture \label{subsubsec:Model_Architecture}}
\begin{figure}[H]
	\centering
	\includegraphics[width=13.5cm]{fig/methods_tcn_arch.pdf}
	\caption{The generic model architecture of the deep learning model to predict initial contacts (ICs) and final contacts (FCs). The inputs are the accelerometer and gyroscope data from a single inertial measurement unit, and are fed to a temporal convolutional network (TCN) (left). The TCN consisted of repeating residual blocks (ResBlock) with exponentially increasing dilation factor (middle). Each ResBlock was built from two sequences of a convolutional layer (Conv), batch normalization layer (BatchNorm), a rectified linear unit activation layer (ReLU), and a dropout layer (DropOut) (right). \label{fig:methods_tcn_arch}}
\end{figure} 
The generic architecture for the deep learning model was based on a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of a sequence of residual blocks with exponentially increasing dilation factor \cite{YuKoltun2016,Bai2018}. Each residual block was built from two sequences of a dilated convolutional layer \cite{YuKoltun2016}, a batch normalization layer \cite{Ioffe2015}, a rectified linear unit (ReLU) activation layer, and dropout layer \cite{Srivastava2014} (Figure~\ref{fig:methods_tcn_arch}). The model was built in Python \cite{VanRossum2009} using the high-level TensorFlow API Keras \cite{Chollet2015,Remy2020}.

For the current analysis only convolutions of the "same" type were considered \cite{Remy2020}, i.e., the model was non-causal and zero-padded to account for edge effects, and the likelihood of an IC or FC was based on input data both before and after the current sample, $n$:
\begin{equation}
	\hat{y}_{i}[n] = f(\cdots,\mathbf{X}[n-1],\mathbf{X}[n],\mathbf{X}[n+1],\cdots), \quad i \in \left\{ \mathrm{IC}, \mathrm{FC}\right\}
	\label{eqn:predictions}
\end{equation}
The number of samples that the predictions at time $n$ "sees", was referred to as the \emph{receptive field} \cite{Lea2017} and was a function of the kernel size and the dilation factors \cite{Remy2020}. Dilation factors were always given as a sequence of increasing powers of 2 \cite{VanDenOord2016,YuKoltun2016,Bai2018}.

The outputs of the TCN block were fed to two separate fully-connected (FCN, or dense) layers, that both were followed by a sigmoid activation layer. Outputs were then predicted separately for ICs and FCs. The mean squared error (MSE) was used as loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

\subsubsection{Hyperparameter Optimization}
In order to find the best model architecture, hyperparameter tuning was perfomed using KerasTuner \cite{OMalley2019}. Here, the number of filters, the kernel size, and the maximum dilation factor (Table \ref{tab:hyperparameter_optimization}) were optimized using a random search strategy \cite{Bergstra2012}.

@@

The model architecture that resulted from the hyperparameter optimization was then trained on the combined set of training and validation data. The trained model was used to predict the likelihoods of ICs and FCs events on the hold-out test set.

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived gait parameters.
\subsubsection{Overall detection performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated evens were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many event that were detected, were actually not annotated (false positives, FP). An annotated event was considered to be detected if any detected event was within 0.075 s. A detected event were defined as peaks in the predicted likelihoods with a minimum height of 0.4, i.e. $40\%$ likely that it corresponded to an event.

From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} \nonumber
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} \nonumber
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \cdot \textrm{precision}}{\textrm{recall} + \textrm{precision}}  \nonumber
	\label{eqn:f1_score}
\end{equation}

\subsubsection{Time Agreement \label{subsubsec:Methods_Time_Agreement}}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), the time agreement between the annotated and detect gait event was given by the time error:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Stride Parameters}
For those trials for which all gait events were detected, and no false positives were detected, the stride time, stance time and swing were calculated. Stride time was the time between two successive ICs of the same foot. Stance time was the time between a FC and the preceding IC of the same foot. Swing time was the time between the IC following the last FC of the same foot.

@@

\subsection{Overall Detection Performance}
The performance of detecting ICs and FCs was objectively quantified by the number of TPs, the number of FNs, and the number of FPs. From these, recall, precision and F$_{1}$ score were calculated (Table \ref{tab:overall_detection_performance}).
\begin{table}[H]
	\caption{Overall detection performance for initial contacts and final contacts as quantified by recall, precision and F$_{1}$ score.\label{tab:overall_detection_performance}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{r|CCCCCC|CCCCCC}
			\toprule
			 & \multicolumn{6}{c}{\textbf{Initial contacts}}	& \multicolumn{6}{c}{\textbf{Final contacts}}\\
			\textbf{Tracked point}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision}	& \textbf{F1}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision} 	& \textbf{F1}\\
			\midrule
			left ankle		& 624	& 19	& 5		& 97\%	& 99\%	& 98\%		& 606	& 32	& 10	& 95\%	& 98\%	& 97\%\\
			right ankle		& 599	& 42	& 8 	& 93\%	& 99\%	& 96\%		& 614	& 17	& 12	& 97\%	& 98\%	& 98\%\\
			left shank		& 605	& 38	& 15	& 94\%	& 98\%	& 96\%		& 585	& 53	& 18	& 92\%	& 97\%	& 94\%\\
			right shank		& 603	& 36	& 15 	& 94\%	& 98\%	& 96\%		& 595	& 30	& 9		& 95\%	& 99\%	& 97\%\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{TP: true positives, FN: false negatives, FP: false positives, F1: F$_{1}$ score.}}
\end{table}

@@

\subsection{Time Agreement}
\begin{figure}[H]
	\centering
	\includegraphics[width=13.5 cm]{fig/box_plots_tracked_points_with_outliers.png}
	\caption{Time errors for initial (left) and final (right) contacts detection, for each of the different tracked points.\label{fig:time_error_box_plots}}
\end{figure}

Time agreement between the annotated and detected events was quantified for the TPs for each of the tracked points (Figure~\ref{fig:time_error_box_plots}). The median time error for each of the tracked points, and for both ICs and FCs was close to zero (Table~\ref{tab:time_errors}), and the largest deviation was -0.005 s, corresponding to one sample period. The IQR was at most 0.020 s, corresponding to 4 sample periods.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CC|CC}
		\toprule
		 & \multicolumn{2}{c}{\textbf{Initial contacts}} & \multicolumn{2}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & \textbf{median} & \textbf{IQR}\\
		 & s & s & s & s \\
		\midrule
		left ankle & 0.000 & 0.020 & 0.000 & 0.010\\
		right ankle & 0.000 & 0.020 & -0.005 & 0.015\\
		left shank & -0.005 & 0.020 & -0.005 & 0.020\\
		right shank & -0.003 & 0.020 & -0.005 & 0.020\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range.}}
\end{table}

\subsection{Stride Parameters}
For those trials for which all gait events were correctly detected (and no false positives were detected), stride time, stance time and swing time were calculated. The mean difference and the limits of agreement between the marker-based annotations and the model-based detections were calculated.

For all stride-specific gait parameters, and for all tracked points, the mean difference was close to zero, i.e., the maximum mean difference was 0.003 s, namely for the calculated swing time of the right ankle (Table~\ref{tab:stride_parameters}). Furthermore, for all gait parameters and for all tracked points the limits of agreement, based on a 95\% confidence interval were distributed around a zero mean difference with the overall limits of agreement at -0.049 s and 0.051 s (Figure~\ref{fig:gait_parameters_bland_altman_plots}).
\begin{table}[H] 
	\caption{Time agreement between the stride-specific parameters. \label{tab:stride_parameters}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{rrCC}
		\toprule
		\textbf{Tracked point} & \textbf{Parameters} & \textbf{Mean difference} & \textbf{Limits of Agreement}\\
		 & & ms & (ms, ms)\\
		\midrule
		left ankle & stride time & 0.001 & (-0.035, 0.036)\\
		 & stance time & 0.002 & (-0.039, 0.042)\\
		 & swing time & -0.001 & (-0.045, 0.043)\\
		\midrule
		right ankle & stride time & 0.000 & (-0.039, 0.040)\\
		 & stance time & -0.002 & (-0.048, 0.044)\\
		 & swing time & 0.003 & (-0.046, 0.051)\\
		\midrule
		left shank & stride time & 0.001 & (-0.039, 0.041)\\
		& stance time & 0.002 & (-0.043, 0.046)\\
		& swing time & -0.001 & (-0.049, 0.047)\\
		\midrule
		right shank & stride time & -0.000 & (-0.031, 0.031)\\
		& stance time & 0.002 & (-0.046, 0.049)\\
		& swing time & -0.002 & (-0.049, 0.046)\\
		\bottomrule
	\end{tabularx}
	\noindent{}
\end{table}

@@

	\bibitem[VanRossum(2009)]{VanRossum2009}
	Van~Rossum,~G.; Drake,~F.L. \textit{Python 3 Reference Manual}; CreateSpace, Scotts Valley, CA, USA, 2009.
	\bibitem[Chollet(2015)]{Chollet2015}
	Chollet,~F. and others. Keras. {\bf 2015}. \url{https://keras.io}.
